Comb associative operads

### 2018-05-30 Article.tex
Modification notation : d changé par des gamma

Let us now define for any $d \geq 1$ the \Def{$d$-comb associative
operad} $\CAs{d}$ as the quotient operad $\Mag/_{\CongrCAs{d}}$ where
$\CongrCAs{d}$ is the smallest operad congruence of $\Mag$
satisfying
\begin{equation} \label{equ:congruence_Asd}
    \underbrace{(\dots (\Product \circ_1 \Product) \circ_1 \dots)
        \circ_1 \Product}
    _{d \mbox{ \footnotesize operands}}
    \enspace \CongrCAs{d} \enspace
    \underbrace{\Product \circ_2
        (\dots \circ_2 (\Product \circ_2 \Product) \dots)}
    _{d \mbox{ \footnotesize operands}}.
\end{equation}
In words, \eqref{equ:congruence_Asd} says that the left and the right
comb binary trees of degree $d$ are equivalent for $\CongrCAs{d}$.
Notice that $\CongrCAs{1}$ is trivial so that $\CAs{1} = \Mag$ and that
$\CongrCAs{2}$ is the operad congruence defined
by~\eqref{equ:congruence_as} so that $\CAs{2} = \As$.

In the sequel we denote by $\LC{d}$ and $\RC{d}$ the left comb and the
right comb of degree $d$, respectively. Hence, $\CongrCAs{d}$ is the
congruence relation induced by $\LC{d}\CongrCAs{d}\RC{d}$.

We denote by CAs the set of $d$-comb associative operads where $d$ belongs
the set of strictly positive integers:
\[\text{CAs}=\left\{\CAs{d}\mid d\geq 1\right\}.\]
Our purpose is to show that CAs admits a lattice structure. For that, we provide
a description of operads morphisms between the elements of CAs.

@@

  Let $d$ and $d'$ be two strictly positive integers. There exists at most
  one operads morphism $\varphi:\CAs{d'}\longrightarrow\CAs{d}$. This morphism
  exists if and only if $\LC{d'}\CongrCAs{d}\RC{d'}$ and it is surjective in this
  case. Moreover, $\varphi$ is injective if and only if $d=d'$, that is if and only
  if $\varphi$ is the identity morphism.

@@

\begin{proof}
  The operad $\CAs{d'}$ is generated by one binary generator $t^{(d')}$, so that
  $\varphi$ is entirely determined by $\varphi\left(t^{(d')}\right)$. Moreover,
  $\varphi\left(t^{(d')}\right)$ has to be of arity $2$ in $\CAs{d}$, so that
  $\varphi\left(t^{(d')}\right)$ has to be equal to $t^{(d)}$. Hence, if $\varphi$
  exists, it is the unique operads morphism from $\CAs{d'}$ and $\CAs{d}$,
  and in this case, $t^{(d)}$ being in the image of $\varphi$, the latter is
  surjective. Moreover, $\varphi$ is well defined if and only if
  $\varphi\left(\LC{d'}\right)$ and $\varphi\left(\RC{d'}\right)$ are equal in
  $\CAs{d}$, that is if and only if $\LC{d'}\CongrCAs{d}\RC{d'}$. In particular,
  if $d'$ is strictly smaller than $d$, then there does not exist any morphism
  $\varphi:\CAs{d'}\longrightarrow\CAs{d}$. Moreover, if $d'$ is strictly smaller
  than $d$, the cardinality of $\CAs{d'}\left(d\right)$ is equal to the cardinality
  of $\CAs{d}(d)$ plus one. Hence, if there exists an injective morphism
  $\varphi:\CAs{d'}\longrightarrow\CAs{d}$, we must have $d=d'$, and in this case,
  $\varphi$ is the identity morphism.
\end{proof}

We define the binary relation $\preceq$ on CAs as follows: we have $\CAs{d}\preceq\CAs{d'}$
if and only if there exists a morphism $\varphi:\CAs{d'}\longrightarrow\CAs{d}$.

\begin{Theorem}\label{thm:poset_CAs}
  The binary relation $\preceq$ is a partially order on \emph{CAs}, that is
  $\left(\emph{CAs},\ \preceq\right)$ is a poset.
  \end{Theorem}

\begin{proof}
  The binary relation $\preceq$ is reflexive since there exists the identity morphism
  of $\CAs{d}$ for every strictly positive integer $d$. It is transitive since if
  there exist two morphismsk  $\varphi:\CAs{d}\longrightarrow\CAs{d'}$ and
  $\psi:\CAs{d''}\longrightarrow\CAs{d'}$, then $\psi\circ\varphi$ is a morphism from
  $\CAs{d''}$ to $\CAs{d}$. Now, let us assume that there exist morphisms
  $\varphi:\CAs{d'}\longrightarrow\CAs{d}$ and $\psi:\CAs{d}\longrightarrow\CAs{d'}$.
  In particular, $\psi\circ\varphi$ and $\varphi\circ\psi$ are endomorphisms of $\CAs{d'}$
  and $\CAs{d}$, respectively. From Proposition~\ref{prop:morphisms_CAs_d}, these two
  morphisms are identity morphisms, so that $\varphi$ and $\psi$ are injective. From
  Proposition~\ref{prop:morphisms_CAs_d}, $d$ and $d'$ are equal, which proves that
  $\preceq$ is anti-symmetric, hence a partial order.
  \end{proof}

@@

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
  Let $d$ be a strictly positive integer and let $t$ and $t'$ be two elements of $\Mag$. If
  $t\CongrCAs{d}t'$, then $\emph{lr}\left(t\right)-\emph{lr}\left(t'\right)$ is divisible by
  $d-1$.
\end{Lemma}

\begin{proof}
  Let $\Rew$ be the rewrite relation on $\Mag$ induced by the rewrite rule
  $\LC{d}\Rew\RC{d}$. In the rest of the proof we fix two elements $t$ and $t'$ of $\Mag$.

  First, we show that $t\Rew t'$ implies that $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$
  is divisible by $d-1$ \textbf{TODO}.

  Assume that $t\overset{*}{\Rew}t'$, so that there exist elements $t_1,\ \cdots,\ t_n$
  of $\Mag$ such that $t_1=t$, $t_n=t'$ and for every $i\in\{1,\cdots,n-1\}$, we have
  $t_i\Rew t_{i+1}$. The integer $\text{lr}\left(t_{i+1}\right)-\text{lr}\left(t_i\right)$
  is divisible by $d-1$ from the first part of the proof, so that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $d-1$.

  Assume that $t\overset{*}{\leftrightarrow}t'$, so that there exist elements $t_1,\ \cdots,\ t_n$
  of $\Mag$ such that $t_1=t$, $t_n=t'$ and for every $i\in\{1,\cdots,n-1\}$, we have
  $t_i\overset{*}{\Rew}t_{i+1}$ or $t_{i+1}\overset{*}{\Rew}t_i$. From the second part of the proof,
  $\text{lr}\left(t_{i+1}\right)-\text{lr}\left(t_i\right)$ is divisible by $d-1$, so that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $d-1$.

  The congruence relation $\CongrCAs{d}$ is the equivalence relation induced by $\Rew$, that
  is we have $t\CongrCAs{d}t'$ if and only if $t\overset{*}{\leftrightarrow}t'$. Hence, from the
  third part of the proof, $t\CongrCAs{d}t'$ implies that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $d-1$.
  \end{proof}

\begin{Proposition} \label{prop:division_CAs}
  Let $d$ and $d'$ be two strictly positive integers. There exists a morphism
  $\varphi:\CAs{d'}\longrightarrow\CAs{d}$ if and only if $\left(d-1\right)\mid\left(d'-1\right)$.
\end{Proposition}
\begin{proof}
  From Proposition~\ref{prop:morphisms_CAs_d}, it is sufficient to show that $\LC{d'}\CongrCAs{d}\RC{d'}$
  if and only if $\left(d-1\right)\mid\left(d'-1\right)$. If $\LC{d'}\CongrCAs{d}\RC{d'}$, from
  Lemma~\ref{lem:left_rank_and_CongrCAs}, $\text{lr}\left(\LC{d'}\right)-\text{lr}\left(\RC{d'}\right)=d'-1$
  is divisible by $d-1$, which shows the direct implication. Conversely, if
  $\left(d-1\right)\mid\left(d'-1\right)$, the rewrite rule $\LC{d}\Rew\RC{d}$ induces the
  following sequence of rewriting steps:
  \[\textbf{Rewriting sequence from left comb of degree d' to right comb of degree d'}\]
  Hence, we have $\LC{d'}\CongrCAs{d}\RC{d'}$.
   \end{proof}

Proposition~\ref{prop:division_CAs} implies the following:
\begin{Theorem}\label{thm:lattice_CAs}
  The poset $\left(\emph{CAs},\ \preceq\right)$ extends into a lattice
  $\left(\emph{CAs},\ \preceq,\ \wedge,\ \vee\right)$, where the lower-
  bound $\wedge$ and the upper-bound $\vee$ are defined as follows:
  \[\CAs{d}\wedge\CAs{d'}=\CAs{\text{gcd}\left(d-1,\ d'-1\right)}\
  \text{and}\ \CAs{d}\vee\CAs{d'}=\CAs{\text{lcm}\left(d-1,\ d'-1\right)}.\]
  \end{Theorem}

>>>>

Let us now define for any $d \geq 1$ the \Def{$\gamma$-comb associative
operad} $\CAs{\gamma}$ as the quotient operad $\Mag/_{\CongrCAs{\gamma}}$ where
$\CongrCAs{\gamma}$ is the smallest operad congruence of $\Mag$
satisfying
\begin{equation} \label{equ:congruence_Asd}
    \underbrace{(\dots (\Product \circ_1 \Product) \circ_1 \dots)
        \circ_1 \Product}
    _{\gamma \mbox{ \footnotesize operands}}
    \enspace \CongrCAs{\gamma} \enspace
    \underbrace{\Product \circ_2
        (\dots \circ_2 (\Product \circ_2 \Product) \dots)}
    _{\gamma \mbox{ \footnotesize operands}}.
\end{equation}
In words, \eqref{equ:congruence_Asd} says that the left and the right
comb binary trees of degree $\gamma$ are equivalent for $\CongrCAs{\gamma}$.
Notice that $\CongrCAs{1}$ is trivial so that $\CAs{1} = \Mag$ and that
$\CongrCAs{2}$ is the operad congruence defined
by~\eqref{equ:congruence_as} so that $\CAs{2} = \As$.

In the sequel we denote by $\LC{\gamma}$ and $\RC{\gamma}$ the left comb and the
right comb of degree $\gamma$, respectively. Hence, $\CongrCAs{\gamma}$ is the
congruence relation induced by $\LC{\gamma}\CongrCAs{\gamma}\RC{\gamma}$.

We denote by CAs the set of $\gamma$-comb associative operads where $\gamma$ belongs
the set of strictly positive integers:
\[\text{CAs}=\left\{\CAs{\gamma}\mid\gamma\geq 1\right\}.\]
Our purpose is to show that CAs admits a lattice structure. For that, we provide
a description of operads morphisms between the elements of CAs.

@@

  Let $\gamma$ and $\gamma'$ be two strictly positive integers. There exists at most
  one operads morphism $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$. This morphism
  exists if and only if $\emph{LC}\left(\gamma'\right)\CongrCAs{\gamma}\emph{RC}\left(\gamma'\right)$ and it is surjective in this
  case. Moreover, $\varphi$ is injective if and only if $\gamma=\gamma'$, that is if and only
  if $\varphi$ is the identity morphism.

@@

\begin{proof}
  The operad $\CAs{\gamma'}$ is generated by one binary generator $t^{(\gamma')}$, so that
  $\varphi$ is entirely determined by $\varphi\left(t^{(\gamma')}\right)$. Moreover,
  $\varphi\left(t^{(\gamma')}\right)$ has to be of arity $2$ in $\CAs{\gamma}$, so that
  $\varphi\left(t^{(\gamma')}\right)$ has to be equal to $t^{(\gamma)}$. Hence, if $\varphi$
  exists, it is the unique operads morphism from $\CAs{\gamma'}$ and $\CAs{\gamma}$,
  and in this case, $t^{(\gamma)}$ being in the image of $\varphi$, the latter is
  surjective. Moreover, $\varphi$ is well defined if and only if
  $\varphi\left(\LC{\gamma'}\right)$ and $\varphi\left(\RC{\gamma'}\right)$ are equal in
  $\CAs{\gamma}$, that is if and only if $\LC{\gamma'}\CongrCAs{\gamma}\RC{\gamma'}$. In particular,
  if $\gamma'$ is strictly smaller than $\gamma$, then there does not exist any morphism
  $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$. Moreover, if $\gamma'$ is strictly smaller
  than $\gamma$, the cardinality of $\CAs{\gamma'}\left(\gamma\right)$ is equal to the cardinality
  of $\CAs{\gamma}(\gamma)$ plus one. Hence, if there exists an injective morphism
  $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$, we must have $\gamma=\gamma'$, and in this case,
  $\varphi$ is the identity morphism.
\end{proof}

We define the binary relation $\preceq$ on CAs as follows: we have $\CAs{\gamma}\preceq\CAs{\gamma'}$
if and only if there exists a morphism $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$.

\begin{Proposition}\label{prop:poset_CAs}
  The binary relation $\preceq$ is a partially order on \emph{CAs}, that is
  $\left(\emph{CAs},\ \preceq\right)$ is a poset.
  \end{Proposition}

\begin{proof}
  The binary relation $\preceq$ is reflexive since there exists the identity morphism
  of $\CAs{\gamma}$ for every strictly positive integer $\gamma$. It is transitive since if
  there exist two morphisms  $\varphi:\CAs{\gamma}\to\CAs{\gamma'}$ and
  $\psi:\CAs{\gamma''}\to\CAs{\gamma'}$, then $\psi\circ\varphi$ is a morphism from
  $\CAs{\gamma''}$ to $\CAs{\gamma}$. Now, let us assume that there exist morphisms
  $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$ and $\psi:\CAs{\gamma}\to\CAs{\gamma'}$.
  In particular, $\psi\circ\varphi$ and $\varphi\circ\psi$ are endomorphisms of $\CAs{\gamma'}$
  and $\CAs{\gamma}$, respectively. From Proposition~\ref{prop:morphisms_CAs_d}, these two
  morphisms are identity morphisms, so that $\varphi$ and $\psi$ are injective. From
  Proposition~\ref{prop:morphisms_CAs_d}, $\gamma$ and $\gamma'$ are equal, which proves that
  $\preceq$ is anti-symmetric, hence a partial order.
  \end{proof}

@@

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
  Let $\gamma$ be a strictly positive integer and let $t$ and $t'$ be two elements of $\Mag$. If
  $t\CongrCAs{\gamma}t'$, then $\emph{lr}\left(t\right)$ and $\emph{lr}\left(t'\right)$ are equal modulo $\gamma-1$: $\emph{lr}\left(t\right)\equiv\emph{lr}\left(t'\right) [\gamma-1]$.
\end{Lemma}

\begin{proof}
  Let $\Rew$ be the rewrite relation on $\Mag$ induced by the rewrite rule
  $\LC{\gamma}\Rew\RC{\gamma}$. In the rest of the proof we fix two elements $t$ and $t'$ of $\Mag$.

  First, we show that $t\Rew t'$ implies that $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$
  is divisible by $\gamma-1$ \textbf{TODO}.

  Assume that $t\overset{*}{\Rew}t'$, so that there exist elements $t_1,\ \cdots,\ t_n$
  of $\Mag$ such that $t_1=t$, $t_n=t'$ and for every $i\in\{1,\cdots,n-1\}$, we have
  $t_i\Rew t_{i+1}$. The integer $\text{lr}\left(t_{i+1}\right)-\text{lr}\left(t_i\right)$
  is divisible by $\gamma-1$ from the first part of the proof, so that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $\gamma-1$.

  Assume that $t\overset{*}{\leftrightarrow}t'$, so that there exist elements $t_1,\ \cdots,\ t_n$
  of $\Mag$ such that $t_1=t$, $t_n=t'$ and for every $i\in\{1,\cdots,n-1\}$, we have
  $t_i\overset{*}{\Rew}t_{i+1}$ or $t_{i+1}\overset{*}{\Rew}t_i$. From the second part of the proof,
  $\text{lr}\left(t_{i+1}\right)-\text{lr}\left(t_i\right)$ is divisible by $\gamma-1$, so that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $\gamma-1$.

  The congruence relation $\CongrCAs{\gamma}$ is the equivalence relation induced by $\Rew$, that
  is we have $t\CongrCAs{\gamma}t'$ if and only if $t\overset{*}{\leftrightarrow}t'$. Hence, from the
  third part of the proof, $t\CongrCAs{\gamma}t'$ implies that
  $\text{lr}\left(t'\right)-\text{lr}\left(t\right)$ is divisible by $\gamma-1$.
  \end{proof}

\begin{Proposition} \label{prop:division_CAs}
  Let $\gamma$ and $\gamma'$ be two strictly positive integers. There exists a morphism
  $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$ if and only if $\left(\gamma-1\right)\mid\left(\gamma'-1\right)$.
\end{Proposition}
\begin{proof}
  From Proposition~\ref{prop:morphisms_CAs_d}, it is sufficient to show that $\LC{\gamma'}\CongrCAs{\gamma}\RC{\gamma'}$
  if and only if $\left(\gamma-1\right)\mid\left(\gamma'-1\right)$. If $\LC{\gamma'}\CongrCAs{d}\RC{\gamma'}$, from
  Lemma~\ref{lem:left_rank_and_CongrCAs}, $\text{lr}\left(\LC{\gamma'}\right)-\text{lr}\left(\RC{\gamma'}\right)=\gamma'-1$
  is divisible by $\gamma-1$, which shows the direct implication. Conversely, if
  $\left(\gamma-1\right)\mid\left(\gamma'-1\right)$, the rewrite rule $\LC{\gamma}\Rew\RC{\gamma}$ induces the
  following sequence of rewriting steps:
  \[\textbf{Rewriting sequence from left comb of degree d' to right comb of degree d'}\]
  Hence, we have $\LC{\gamma'}\CongrCAs{\gamma}\RC{\gamma'}$.
   \end{proof}

Proposition~\ref{prop:division_CAs} implies the following:
\begin{Theorem}\label{thm:lattice_CAs}
  The poset $\left(\emph{CAs},\ \preceq\right)$ extends into a lattice
  $\left(\emph{CAs},\ \preceq,\ \wedge,\ \vee\right)$, where the lower-
  bound $\wedge$ and the upper-bound $\vee$ are defined as follows:
  \[\CAs{\gamma}\wedge\CAs{\gamma'}=\CAs{\text{gcd}\left(\gamma-1,\ \gamma'-1\right)}\
  \text{and}\ \CAs{\gamma}\vee\CAs{\gamma'}=\CAs{\text{lcm}\left(\gamma-1,\ \gamma'-1\right)}.\]
  \end{Theorem}
